---
title: "Case Study: How Does a Bike-Share Navigate Speedy Success?"
author: "Athena Pelfrey"
date: "June 23, 2023"
output: html_notebook
---

```{r importing packages, include=FALSE}

# Clear workspace
rm(list = ls()) 

library(tidyverse)
library(lubridate)
library(ggplot2)
library(mosaic)
library(presenter)
```



# Introduction

This case study was based on an assignment in the Coursera course *Google Data Analytics Capstone: A Complete Case Study*. 

**About the Company: ** In 2016, Cyclistic launched a successful bike-share offering. Since then, the program has grown to a fleet of 5,824 bicycles that
are geotracked and locked into a network of 692 stations across Chicago. The bikes can be unlocked from one station and
returned to any other station in the system anytime.

**Company Goal:** Design marketing strategies aimed at converting casual riders into annual members.

**Questions:** Three questions will guide the future marketing program:

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

**Task:** Produce a report with the following deliverables to answer the first question.

1. A clear statement of the business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings
6. Your top three recommendations based on your analysis


# Data Importing and Cleaning
Steps:

- Import data for 2019
- Rename columns for consistency
- Change data types for consistency
- Combine datasets to a single dataframe
- Add new columns for month, day, hour, and day_of_week
- Filter out trips longer than 24 hours

Here is a quick look at this dataset.
```{r data wrangling, include=FALSE, message=FALSE, warning=FALSE}
# Import data
q1_2019 <- read_csv("/Users/aspelfrey/Desktop/CaseStudyCoursera/Data/Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("/Users/aspelfrey/Desktop/CaseStudyCoursera/Data/Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("/Users/aspelfrey/Desktop/CaseStudyCoursera/Data/Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("/Users/aspelfrey/Desktop/CaseStudyCoursera/Data/Divvy_Trips_2019_Q4.csv")

# Rename columns for consistency.
q1_2019 <- rename(q1_2019
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,trip_duration = tripduration
                  ,bike_id = bikeid
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name
                  ,end_station_id = to_station_id 
                  ,user_type = usertype
                  ,birth_year = birthyear
                  )

q3_2019 <- rename(q3_2019
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,trip_duration = tripduration
                  ,bike_id = bikeid
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name
                  ,end_station_id = to_station_id 
                  ,user_type = usertype
                  ,birth_year = birthyear
                  )
q4_2019 <- rename(q4_2019
                  ,started_at = start_time  
                  ,ended_at = end_time  
                  ,trip_duration = tripduration
                  ,bike_id = bikeid
                  ,start_station_name = from_station_name 
                  ,start_station_id = from_station_id 
                  ,end_station_name = to_station_name
                  ,end_station_id = to_station_id 
                  ,user_type = usertype
                  ,birth_year = birthyear
                  )

q2_2019 <- rename(q2_2019
                  ,trip_id = "01 - Rental Details Rental ID"
                  ,started_at = "01 - Rental Details Local Start Time"
                  ,ended_at = "01 - Rental Details Local End Time"  
                  ,trip_duration = "01 - Rental Details Duration In Seconds Uncapped"
                  ,bike_id = "01 - Rental Details Bike ID" 
                  ,start_station_name = "03 - Rental Start Station Name"
                  ,start_station_id = "03 - Rental Start Station ID" 
                  ,end_station_name = "02 - Rental End Station Name" 
                  ,end_station_id = "02 - Rental End Station ID" 
                  ,user_type = "User Type"
                  ,gender = "Member Gender" 
                  ,birth_year = "05 - Member Details Member Birthday Year" 
                  )

# Change data types for consistency
q1_2019 <-  mutate(q1_2019, trip_id = as.character(trip_id)
                   ,bike_id = as.character(bike_id))
q2_2019 <-  mutate(q2_2019, trip_id = as.character(trip_id)
                   ,bike_id = as.character(bike_id))
q3_2019 <-  mutate(q3_2019, trip_id = as.character(trip_id)
                   ,bike_id = as.character(bike_id))
q4_2019 <-  mutate(q4_2019, trip_id = as.character(trip_id)
                   ,bike_id = as.character(bike_id))

```

```{r combine datasets, echo=FALSE}
# Combine datasets

all_trips <- bind_rows(q1_2019, q2_2019, q3_2019, q4_2019)

# Rename user types
all_trips <- all_trips |> 
  mutate(user_type = 
           case_when(user_type == 'Subscriber' ~ 'Member',
                     user_type == 'Customer' ~ 'Casual'),
         trip_duration = hms::as_hms(trip_duration))
all_trips <- all_trips |>
  mutate(year = format(as.Date(started_at),"%Y")) |>
  mutate(month = format(as.Date(started_at),"%m")) |>
  mutate(day = format(as.Date(started_at),"%d")) |>
  mutate(hour = format(as.POSIXct(started_at),"%H")) |>
  mutate(day_of_week = format(as.Date(started_at),"%A"))

head(all_trips)
```

Quick summaries of data.
```{r echo=FALSE}
summary(all_trips)
```


Retrieve statistics for **trip duration** grouped by user type.
```{r echo=FALSE}
all_trips %>%
   group_by(user_type) %>%
   summarise("count" = n(),
             "mean" = hms::as_hms(mean(trip_duration)), 
             "median" = hms::as_hms(median(trip_duration)),
             "min" = hms::as_hms(min(trip_duration)),
             "max" = hms::as_hms(max(trip_duration))) %>%
  pivot_summary()
```
Filter out all trips lasting longer than 24 hours. 
```{r echo=FALSE}
n_mem_filtered <- sum(all_trips_v2$user_type=='Member')
n_cas_filtered <- sum(all_trips_v2$user_type=='Casual')
n_mem_all <- sum(all_trips$user_type=='Member')
n_cas_all <- sum(all_trips$user_type=='Casual')

col_n_filtered <- c(n_mem_filtered, n_cas_filtered)
col_n_all <- c(n_mem_all, n_cas_all)
col_pct_filtered <- col_n_filtered/col_n_all

data.frame("Total Trips" = col_n_all, "Trips less 24 hrs" = col_n_filtered, "Percentage" = col_pct_filtered)
```
Summarize Trip Duration for Filtered data
```{r echo=FALSE}
all_trips_v2 <- filter(all_trips, trip_duration<=hms::as_hms('24:00:00'))

all_trips_v2 %>%
   group_by(user_type) %>%
   summarise("count" = n(),
             "mean" = hms::as_hms(mean(trip_duration)), 
             "median" = hms::as_hms(median(trip_duration)),
             "min" = hms::as_hms(min(trip_duration)),
             "max" = hms::as_hms(max(trip_duration))) %>%
  pivot_summary()
```


# Data Analysis: Day of Week

Retrieve statistics for **number of rides** and **trip duration** grouped by **user type** and **day of week**.
```{r echo=FALSE}
options(dplyr.summarise.inform = FALSE)

level_order = c("Sunday", "Monday", "Tuesday", "Wednesday","Thursday", "Friday", "Saturday")

trips_by_day_of_week <- all_trips_v2 %>%
   group_by(day_of_week, user_type) %>%
   summarise("number_of_rides" = n(),
     "mean" = hms::as_hms(mean(trip_duration)), 
     "median" = hms::as_hms(median(trip_duration)),
     "min" = hms::as_hms(min(trip_duration)),
     "max" = hms::as_hms(max(trip_duration))) %>%
  arrange(factor(day_of_week, levels = level_order), user_type)
trips_by_day_of_week
```

```{r echo=FALSE}
ggplot(trips_by_day_of_week,aes(x = day_of_week, y = number_of_rides, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = '', y = 'Number of Rides') +
  scale_x_discrete(limits =level_order)

ggplot(trips_by_day_of_week,aes(x = day_of_week, y = mean, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = '', y = 'Average Trip Duration') +
  scale_x_discrete(limits =level_order)
```


# Data Analysis: Time of Year

Retrieve statistics for **number of rides** and **trip duration** grouped by **user type** and **month**.
```{r echo=FALSE}
trips_by_month <- all_trips_v2 %>%
   group_by(month, user_type) %>%
   summarise("number_of_rides" = n(),
     "mean" = hms::as_hms(mean(trip_duration)), 
     "median" = hms::as_hms(median(trip_duration)),
     "min" = hms::as_hms(min(trip_duration)),
     "max" = hms::as_hms(max(trip_duration)))
trips_by_month
```

```{r echo=FALSE}
ggplot(trips_by_month,aes(x = month, y = number_of_rides, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = 'Month', y = 'Number of Rides')

ggplot(trips_by_month,aes(x = month, y = mean, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = 'Month', y = 'Average Trip Duration')
  
```

# Data Analysis: Time of Day

Retrieve statistics for **number of rides** and **trip duration** grouped by **user type** and **hour of the day**.

```{r echo=FALSE}
trips_by_time_of_day <- all_trips_v2 |> 
  group_by(hour, user_type) |>
  summarise("number_of_rides" = n(),
     "mean" = hms::as_hms(mean(trip_duration)), 
     "median" = hms::as_hms(median(trip_duration)),
     "min" = hms::as_hms(min(trip_duration)),
     "max" = hms::as_hms(max(trip_duration)))
trips_by_time_of_day
```

```{r echo=FALSE}
ggplot(trips_by_time_of_day,aes(x = hour
                                , y = number_of_rides
                                , fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = 'Hour of Day', y = 'Number of Rides') + 
  annotate(geom = "text", x = 8, y = 300000, label='8 AM') + 
  annotate(geom = "text", x = 17, y = 400000, label='5 PM')

ggplot(trips_by_time_of_day,aes(x = hour, y = mean, fill = user_type)) +
  geom_col(position = "dodge") +
  labs(x = 'Hour of Day', y = 'Average Trip Duration')
```